They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-07 presentation.qmd
move vectorized_prop_test, p_cod_round_win_prop_after_x_wins_y_rounds

@@

```{r}
#| label: cod_series_outcomes_chi

cod_rounds_per_series <- cod_streaks_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )
avg_actual_cod_rounds_per_series <- round(cod_rounds_per_series$actual, 2)
avg_expected_cod_rounds_per_series <- round(cod_rounds_per_series$expected, 2)

@@

## This adjustment is something that is not done in the paper. I do it here to reduce
##   the magnitude of the exponents.
cod_pm <- pmin(cod_n6, cod_n7, cod_n8, cod_n9, cod_n10, cod_n11)
cod_n6adj <- round(10 * cod_n6 / cod_pm)
cod_n7adj <- round(10 * cod_n7 / cod_pm)
cod_n8adj <- round(10 * cod_n8 / cod_)
cod_n9adj <- round(10 * cod_n9 / cod_pm)
cod_n10adj <- round(10 * cod_n10 / cod_pm)
cod_n11adj <- round(10 * cod_n11 / cod_pm)

@@

@@

```{r}
#| label: qs-funcs
#| eval: true
load_qs <- function(name) {
  res <- qs::qread(sprintf('data/%s.qs', name))
  assign(name, value = res, envir = .GlobalEnv)
}
save_qs <- function(df, name = deparse(substitute(df))) {
  qs::qsave(df, sprintf('data/%s.qs', name))
}

@@

summarize_cod_win_prop_after_x_wins <- function(k) {
  stopifnot((k >= 2) & (k <= 5))

  probs_nx_k <- tibble(
    n_rounds = 7L:11L
  ) |> 
    mutate(
      prior_p = 6L / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  cols <- sprintf('won_prior_round%d', 2:k)
  col_syms <- syms(cols)
  cod_round_streaks_after_x <- cod_round_streaks |> 
    drop_na(!!!col_syms) |> 
    count(n_rounds, win_round, won_prior_round1, !!!col_syms, sort = TRUE)

  cod_round_win_prop_after_x_wins <- cod_round_streaks_after_x |> 
    filter(won_prior_round1, !!!col_syms)

  cod_round_win_prop_after_x_wins |> 
    group_by(n_rounds) |> 
    mutate(
      total = sum(n),
      prop = n / total
    ) |> 
    ungroup() |> 
    filter(win_round) |> 
    select(n_rounds, n, total, prop) |> 
    arrange(n_rounds) |> 
    inner_join(
      probs_nx_k,
      by = 'n_rounds'
    ) |> 
    nest(data = -c(n_rounds)) |> 
    mutate(
      gt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'greater') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      lt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'less') |> 
          broom::tidy() |>
          pull(p.value)
      )
    ) |> 
    unnest(data)
}

@@

```{r}
#| label: cod_round_win_prop_after_x_wins-eval
#| eval: true
load_qs('cod_round_win_prop_after_x_wins')
cod_round_win_prop_after_x_wins |> 
  transmute(
    `Win Streak` = win_streak,
    `Rounds` = n_rounds,
    `N` = scales::comma(total),
    across(c(prop, expected_poststreak_p), ~scales::percent(.x, accuracy = 0.1)),
    across(
      ends_with('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  rename(
    `Actual post-streak win %` = prop,
    `Expected post-streak win %` = expected_poststreak_p,
    `Binom. > p-value` = gt_p_value,
    `Binom. < p-value` = lt_p_value
  ) |> 
  knitr::kable()

@@

```{r}
#| label: p_cod_o_win_r1_vs_post_round_prop
# https://raw.githubusercontent.com/dgrtwo/splittestr/master/R/vectorized-prop-test.R
vectorized_prop_test_approx <- function(a, b, c, d) {
  n1 <- a + b
  n2 <- c + d
  n <- n1 + n2
  p <- (a + c) / n
  E <- cbind(p * n1, (1 - p) * n1, p * n2, (1 - p) * n2)

calculate_map_o_round_win_prop <- function(df) {
  df |> 
    count(map, is_offense, win_round) |> 
    group_by(map, is_offense) |> 
    mutate(total = sum(n), prop = n / total) |> 
    ungroup() |>
    filter(win_round, is_offense) |>
    select(
      map,
      rounds_won = n,
      total_rounds = total,
      win_round_prop = prop
    ) |>
    arrange(desc(win_round_prop))
}

>>>>

# https://raw.githubusercontent.com/dgrtwo/splittestr/master/R/vectorized-prop-test.R
vectorized_prop_test_approx <- function(a, b, c, d) {
  n1 <- a + b
  n2 <- c + d
  n <- n1 + n2
  p <- (a + c) / n
  E <- cbind(p * n1, (1 - p) * n1, p * n2, (1 - p) * n2)

@@

```{r}
#| label: cod_series_outcomes_chi
cod_rounds_per_series <- cod_streaks_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )
avg_actual_cod_rounds_per_series <- round(cod_rounds_per_series$actual, 2)
avg_expected_cod_rounds_per_series <- round(cod_rounds_per_series$expected, 2)

@@

## This adjustment is something that is not done in the paper. I do it here to reduce
##   the magnitude of the exponents.
cod_pm <- pmin(cod_n6, cod_n7, cod_n8, cod_n9, cod_n10, cod_n11)
cod_n6adj <- round(10 * cod_n6 / cod_pm)
cod_n7adj <- round(10 * cod_n7 / cod_pm)
cod_n8adj <- round(10 * cod_n8 / cod_pm)
cod_n9adj <- round(10 * cod_n9 / cod_pm)
cod_n10adj <- round(10 * cod_n10 / cod_pm)
cod_n11adj <- round(10 * cod_n11 / cod_pm)

@@

There's a clear bias towards shorter series than expected.

@@

```{r}
#| label: qs-funcs
#| eval: true
load_qs <- function(name) {
  res <- qs::qread(sprintf('data/%s.qs', name))
  assign(name, value = res, envir = .GlobalEnv)
}
save_qs <- function(df, name = deparse(substitute(df))) {
  qs::qsave(df, sprintf('data/%s.qs', name))
}
```

@@

.postprocess_round_streaks <- function(k, round_streaks, probs_nx_k, ...) {

  cols <- sprintf('won_prior_round%d', 2:k)
  col_syms <- syms(cols)
  round_streaks_after_x <- round_streaks |> 
    drop_na(!!!col_syms) |> 
    count(n_rounds, ..., win_round, won_prior_round1, !!!col_syms, sort = TRUE)

  round_win_prop_after_x_wins <- round_streaks_after_x |> 
    filter(won_prior_round1, !!!col_syms)

  round_win_prop_after_x_wins |> 
    group_by(n_rounds, ...) |> 
    mutate(
      total = sum(n),
      prop = n / total
    ) |> 
    ungroup() |> 
    filter(win_round) |> 
    select(n_rounds, ..., n, total, prop) |> 
    arrange(n_rounds, ...) |> 
    inner_join(
      probs_nx_k
    ) |> 
    nest(data = -c(n_rounds, ...)) |> 
    mutate(
      gt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'greater') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      lt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'less') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      gt_is_signif = gt_p_value <= 0.05,
      lt_is_signif = lt_p_value <= 0.05,
      is_signif = case_when(
        gt_is_signif ~ 'Better',
        lt_is_signif ~ 'Worse',
        TRUE ~ 'Neither'
      )
    ) |> 
    unnest(data)
}

summarize_cod_win_prop_after_x_wins <- function(k) {
  probs_nx_k <- tibble(
    n_rounds = 7L:11L
  ) |> 
    mutate(
      prior_p = 6L / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  .postprocess_round_streaks(
    k = k,
    round_streaks = cod_round_streaks,
    probs_nx_k = probs_nx_k
  )
}

@@

```{r}
#| label: cod_round_win_prop_after_x_wins_and_rounds
summarize_cod_win_prop_after_x_wins_y_rounds <- function(k) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    round = 3L:11L
  ) |> 
    filter(round >= (k + 1L)) |> 
    mutate(
      prior_p = 6L / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  .postprocess_round_streaks(
    k = k,
    round_streaks = cod_round_streaks,
    probs_nx_k = probs_nx_k,
    round
  )
}

cod_round_win_prop_after_x_wins_and_rounds <- tibble(win_streak = 2L:4L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_x_wins_and_rounds)
  ) |>
  unnest(data)
save_qs(cod_round_win_prop_after_x_wins_and_rounds)
```

```{r}
#| label: cod_round_win_prop_after_x_wins-table
#| eval: false
load_qs('cod_round_win_prop_after_x_wins')
cod_round_win_prop_after_x_wins |> 
  transmute(
    `Win Streak` = win_streak,
    `Rounds` = n_rounds,
    `N` = scales::comma(total),
    across(c(prop, expected_poststreak_p), ~scales::percent(.x, accuracy = 0.1)),
    across(
      ends_with('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  rename(
    `Actual post-streak win %` = prop,
    `Expected post-streak win %` = expected_poststreak_p,
    `Binom. > p-value` = gt_p_value,
    `Binom. < p-value` = lt_p_value
  ) |> 
  knitr::kable()
```

```{r}
#| label: cod_round_win_prop_after_x_wins-plot
load_qs('cod_round_win_prop_after_x_wins')

ggsci_grey <- extract_ggsci_mid_color('grey')
p_cod_round_win_prop_after_x_wins <- cod_round_win_prop_after_x_wins |> 
  ggplot() +
  geom_tile(
    color = blackish_background,
    height = 0.8,
    aes(
      fill = lt_is_signif,
      y = win_streak + 0.5, 
      x = n_rounds + 0.5
    )
  ) +
  geom_richtext(
    fill = NA_character_,
    label.color = NA_character_,
    family = font,
    color = blackish_background,
    size = 11 / .pt,
    fontface = 'bold',
    aes(
      label = sprintf(
        '%s %s <span style="color:white">%s</span><br/>(%s)',
        scales::percent(prop, accuracy = 1), 
        case_when(
          round(prop, 1) == round(expected_poststreak_p, 1) ~ '=',
          prop >= expected_poststreak_p ~ '>=', 
          TRUE ~ '<'
        ),
        scales::percent(expected_poststreak_p, accuracy = 1),
        scales::comma(n)
      ),
      y = win_streak + 0.5, 
      x = n_rounds + 0.5
    )
  ) +
  guides(
    fill = guide_legend(
      'Significantly less than expected?'
    )
  ) +
  scale_y_continuous(
    labels = 2:4,
    breaks = seq(2.5, 4.5, by = 1),
    expand = c(0, 0.5)
  ) +
  scale_x_continuous(
    labels = 7:11,
    breaks = seq(7.5, 11.5, by = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c(ggsci_grey, ggsci_cyan),
    labels = c('No', 'Yes')
  ) +
  theme(
    legend.position = 'top',
    legend.text = element_text(color = 'white', size = 12),
    panel.grid.major = element_blank(),
    axis.title = element_text(hjust = 0.5),
    plot.title = ggtext::element_markdown(hjust = 0.5),
    axis.text = element_text(size = 16, face = 'bold')
  ) +
  common_cod_labs() +
  labs(
    title = 'Post win streak round win %',
    y = 'Win streak length',
    x = '# of rounds played in series'
  ) +
  geom_curve(
    data = data.frame(),
    color = 'white',
    arrow = arw,
    curvature = 0.1,
    aes(
      x = 7.8,
      y = 5.11,
      xend = 7.7,
      yend = 4.8
    )
  ) +
  geom_text(
    data = data.frame(),
    hjust = 0,
    vjust = 0.5,
    color = 'white',
    size = 12 / .pt,
    aes(
      label = 'Expected %',
      x = 7.81,
      y = 5.12
    )
  ) +
  geom_curve(
    data = data.frame(),
    color = 'white',
    arrow = arw,
    curvature = 0.25,
    aes(
      x = 7.8,
      y = 5.41,
      xend = 7.3,
      yend = 4.8
    )
  ) +
  geom_text(
    data = data.frame(),
    hjust = 0,
    vjust = 0.5,
    color = 'white',
    size = 12 / .pt,
    aes(
      label = 'Actual %',
      x = 7.81,
      y = 5.42
    )
  )
p_cod_round_win_prop_after_x_wins
save_plot_for_slide(p_cod_round_win_prop_after_x_wins)
```

::: fragment
![](plots/cod_round_win_prop_after_x_wins.png)

-   There is no evidence for momentum, for both streaks of 2 and 3. In fact, for series that go to 9, 10, or 11 rounds, there is evidence that the streaking teams perform significantly worse than expected in the post-streak round. And for streaks of 4, teams perform worse than expected in series ending in 7, 8, 10, and 11 rounds.
:::

```{r}
load_qs('cod_round_win_prop_after_x_wins')
ggsci_orange <- extract_ggsci_mid_color('orange')
p_cod_round_win_prop_after_x_wins_y_rounds <- cod_round_win_prop_after_x_wins_y_rounds |> 
  mutate(
    is_signif = case_when(
      gt_is_signif ~ 'Better',
      lt_is_signif ~ 'Worse',
      TRUE ~ 'Neither'
    )
  ) |> 
  # filter(win_streak %in% c(3, 4)) |> 
  # filter(n >= 20) |> 
  ggplot() +
  geom_tile(
    color = blackish_background,
    height = 0.8,
    aes(
      fill = is_signif,
      x = round + 0.5, 
      y = n_rounds + 0.5
    )
  ) +
  geom_richtext(
    fill = NA_character_,
    label.color = NA_character_,
    family = font,
    color = blackish_background,
    size = 11 / .pt,
    fontface = 'bold',
    aes(
      label = sprintf(
        '%s %s <span style="color:white">%s</span><br/>(%s)',
        scales::percent(prop, accuracy = 1), 
        case_when(
          round(prop, 1) == round(expected_poststreak_p, 1) ~ '=',
          prop >= expected_poststreak_p ~ '>=', 
          TRUE ~ '<'
        ),
        scales::percent(expected_poststreak_p, accuracy = 1),
        scales::comma(n)
      ),
      x = round + 0.5, 
      y = n_rounds + 0.5
    )
  ) +
  guides(
    fill = guide_legend(
      'Significantly better/worse?'
    )
  ) +
  scale_x_continuous(
    labels = 4:11,
    breaks = seq(4.5, 11.5, by = 1),
    expand = c(0, 0.5)
  ) +
  scale_y_continuous(
    labels = 7:11,
    breaks = seq(7.5, 11.5, by = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c(ggsci_cyan, ggsci_grey, ggsci_orange),
    labels = c('Better', 'Neither', 'Worse')
  ) +
  theme(
    legend.position = 'top',
    legend.text = element_text(color = 'white', size = 12),
    panel.grid.major = element_blank(),
    axis.title = element_text(hjust = 0.5),
    plot.title = ggtext::element_markdown(hjust = 0.5),
    axis.text = element_text(size = 16, face = 'bold')
  ) +
  common_cod_labs() +
  labs(
    title = 'Post win streak round win %',
    x = 'Round',
    y = '# of rounds played in series'
  ) +
  facet_wrap(~win_streak)
p_cod_round_win_prop_after_x_wins_y_rounds
```

@@

```{r}
#| label: p_cod_o_win_r1_vs_post_round_prop
calculate_map_o_round_win_prop <- function(df) {
  df |> 
    count(map, is_offense, win_round) |> 
    group_by(map, is_offense) |> 
    mutate(total = sum(n), prop = n / total) |> 
    ungroup() |>
    filter(win_round, is_offense) |>
    select(
      map,
      rounds_won = n,
      total_rounds = total,
      win_round_prop = prop
    ) |>
    arrange(desc(win_round_prop))
}
